|

### 2020-11-27 paper/v1/broadband.speed.covid.Rmd
after conflicts resolved, addition in intro, lit rev, methods and data.

---
title: |
  WFH and broadband speed (title needs rework)
type: ARTICLE TEMPLATE
author:
  - name: A. N. Author
    affil: a
    email: latex.helpdesk@tandf.co.uk
  - name: John Smith
    affil: b
    email: john.smith@uni-freiburg.de
affiliation:
  - num: a
    address: |
      Taylor & Francis, 4 Park Square, Milton Park, Abingdon, UK
  - num: b
    address: |
      Institut für Informatik, Albert-Ludwigs-Universität, Freiburg, Germany
bibliography: bibliography.bib
appendix: appendix.tex
abstract: |
  TBC
keywords: |
  covid; internet; working from home; broadband speed; time series clusters
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

@@

- [our google doc](https://docs.google.com/document/d/1PWjkmgzWGYKR9wFogKYw7l-8mZLoORt593x-Tu-f2-M/edit#heading=h.i5om1o8wpcd9)

During the pandemic, working from home using digital technologies, whether partially 
or exclusively, was transformed from a niche means of accessing work, albeit one that 
had been on a slow, upward trend, to a widespread way of life in many countries.
The ability to work from home or telecommute meant millions retained their jobs and,
to a varying extent, maintained productivity during periods of strict
lockdown around the world.
However, this ability has not been evenly distributed socially or spatially,
creating a new type of digital divide.
On one side are those who can work from home, supported by digital technologies, 
and have thus been able to enjoy both economic resilience and greater personal safety.
On the other side, previously employed individuals have been forced to accept
furlough or redundancy packages unless they are part of the cadre of essential
workers, who are potentially at high risk of infection.
Whilst the basis for this new digital divide can be viewed as mainly occupational, 
we consider whether the divide is also technological.

<!-- our approach -->
Using the UK as a case study, this paper aims to understand how the resilience
of broadband speeds in terms of both quality and reliability of service may reinforce
or redress the spatial and social dimensions of the digital division exposed by the
pandemic.
To do so, we employ volunteered geographic data on individual broadband speed tests
and state-of-the-art time-series clustering methods to create clusters of UK local
authorities with similar temporal signatures of experienced internet speeds.
We then associate these clusters of local authorities with their socioeconomic
and geographic characteristics to explore how they overlap with or diverge from
the existing economic and digital geography of the UK. **need to add urban-rural to analysis**
Our analysis enables us to better understand how the spatial and social distribution 
of both occupations and online accessibility intersect to enable or hinder the 
practice of telecommuting at a time of extreme demand. 
We will also consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a common means of accessing work, and broadband
services and infrastructure must be fit for purpose.
**LET'S LEAVE IT FOR NOW, BUT I THINK WE CAN CRYSTALISE MORE THE RQ**

<!-- capability to WFH -->
The capability to work from home has previously been studied from the
perspective of whether work tasks in a given occupation both can be and are
allowed to be performed using digital technologies independently of 
location or co-location with colleagues, including supervisors 
[@allen2015effective; @singh2013modeling].
However, successful telecommuting also requires that the quality and
reliability of digital services, particularly home internet connection speeds,
enable the completion of work tasks with a minimum of delay or interruption.
Prior to the pandemic, the performance of broadband services with respect to
telecommuters was never tested at scale, as working from home and connecting to
colleagues and workplace resources via broadband was the purview of a small minority 
of workers. 
Instead, leisure use in the evening, when video streaming services are at their peak, 
has been used to benchmark broadband performance and service delivery by different 
Internet Service Providers (ISPs), at least in the UK [@ofcom2017].
Yet the shift towards telecommuting during various stages of lockdown around the world
has been drastic and there are speculations that post-Covid, the tendency to work from
home will be much higher than pre-Covid, raising questions around whether internet 
services can accommodate the increased demand.
For example, $47$% of people in employment in the UK worked soley from home in April
$2020$, whilst the same figure only reached $5$% the year before [@ons2020; 
@ons2020lm2019].
A back of the envelope calculation suggests that up to 40% of the working force
could work from home on an ongoing basis [@batty2020editorial].
Similar figures have been reported for other countries [@felstead2020homeworking].
For instance, $37$% of the European workforce worked from home in April $2020$ with
countries like Finland reaching $60$% [@eurofound2020].
In the US, almost half of the working population worked from home during the same
period because of the pandemic [@brynjolfsson2020covid], 
and a recent estimate indicated that $37$% of all jobs in the US can be permanently 
performed entirely from home [@NBERw26948].

<!-- ICT infrastructure and digital divides-->
None of these changes could have happened in the absence of reliable ICT
infrastructure -- both in terms of software and hardware.
But while software innovations are easily diffused across space and
society^[See for example the huge success of videoconferencing apps
[@marks2020zoom].],
the same does not apply for ICT hardware infrastructure such as internet
broadband connectivity.
The literature describes first level digital divides in terms of the availability and
quality of internet connectivity, such as that manifest in different geographies 
in the UK [@riddlesden2014broadband; @philip2017digital]. 
Second level digital divides consider the presence or lack of the necessary skills 
to effectively utilise digital technologies and the internet
[@blank2014dimensions; @van2011internet].
The third level focuses on the differentiated returns of internet use **for...? add a phrase here introducing economic resilience?**
[@stern2009levels; @van2014digital; @van2015third].
The capability to telecommute is related to all three levels of digital divides,
but more importantly leads to differentiated outcomes regarding the economic 
resilience of people and places during the current pandemic.
We identify this as a new digital divide, one that fundamentally alters the 
potential returns of internet use for the user and wider community, 
assumes skills or functions that are present in some occupations but not in others, 
and relies upon access to high quality internet services.
As the quality of internet infrastructure and services, as well as variation in 
occupations are spatially dependent and clustered in space, our approach offers a 
framework for understanding which types of places,  
are more likely to land on the right side of this new digital divide. 
By asking how resilient broadband speeds are as experienced in different parts of the 
UK during a time of extreme demand, 
we interrogate which places benefit from greater the economic resilience 
digital technologies can offer, not only during the pandemic, 
but also into the future.

The structure of this paper is as follows.
First we will review the literature on telecommuting and digital divides to better 
understand the origins of the new digital divide revealed by the pandemic and its
impact on the economic resilience of different places.
We then describe our data and methodology.
Our results section first describes the how internet services vary in the 
different clusters and then assesses whether these clusters replicate or repudiate
other socio-economic and geographic patterns of economic resilience.

@@

## from telecommuting to #WFH

@@

## digital divides and economic resilience
<!-- WFH and digital divides-->
Our understanding of telecommuting as a product of enabled occupations can be 
described as a manifestation of the third level digital divide, as those who are able 
to use digital technologies to work from home benefit from a high rate of return on 
their use of the internet in terms of autonomy, flexibility, and time saved from 
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
These returns have been even greater during the Covid-19 crisis, 
when those with the ability to telecommute also have the ability to maintain 
their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the first 
level digital divide, which is associated with access and quality of internet 
connectivity.
For example, the systematic review from @SALEMINK2017360 highlights the 
infrastructure quality differences between urban and rural areas 
in various advanced economies.
Whether this variation in infrastructure quality affects the spatial footprint of 
telecommuting has not previously been measured, although there are indications that
those who purchase high speed connections consume more data of all sorts and use their
connections for a variety of purposes [@hauge2011consumer].
Another study identifies a correlation between access to internet services and a 
reduction in household transport spend across 33 coutries [@bris2017ict]. 
Whether the implication is less travel because of increased telecommuting, 
or whether internet access enables more efficient travel, this finding is 
an indication of the potential household savings better internet services offer.
Thus, the extreme demand during the pandemic provides a new opportunity to 
understand how infrastructure accessibility, quality, and reliability 
affects telecommuting, particularly as working from home during the pandemic required
high volumes of bandwidth-intensive video conferencing in order to avoid the 
face-to-face contact that could increase the spread of infection.
By first answering questions about internet service resilience, we can also refine our
understanding of how this has contributed to or reduced the new digital divide, 
where economic resilience has been dependent upon the capability to work from home.

## covid and working from home, cities, urban structure
This is aligned with the regional economic resilience literature, which
underlines the differentiated capacity of cities and regions to escape or
recover from economic crises [@martin2012regional; @kitsos2018economic]. 
Different cities have different industrial and occupational profiles, 
and these will affect the potential for telecommuting... 
**Fill in more here - relate to 2nd and 3rd level digital divide, economic geography of UK**

## other subsection?
- circumstances where work can be carried out more flexibly in space and time
- how relate to digital divides
- capability theory of mobility?
- broadband tech stuff? broadband studies / divides / resilience

@@

# Data and descriptive statistics

<!-- PARA introduce broadband speeds -->
This paper assesses the quality and reliability of upload speeds by time of day and 
day of the week across local authorities in the UK during the time when the population
were told to work from home if at all possible.
The experience of upload speeds is not the same as the maximum speeds offered by an 
ISP or possible speeds over a particular type and length of connection. 
We use volunteered geographic information in the form of speed-checks run by users to 
test their experienced broadband speeds, upload and download. 
This primary data was provided by Speedchecker Ltd, 
a private company that allows internet users to check their own broadband upload / 
download speeds, and stores every speed-check with a timestamp and geolocation.
Since the quality and reliability of internet services vary in time and space due to 
both supply and demand-side influences, we use a number of different measures of 
experienced upload speeds. 
These include: a) mean, experienced connection speed, 
b) standard deviation or the amount of fluctuation from the mean, and 
c) the variation in speeds at particular times of day when working from home is more 
likely to take place. 
We take account of all three measurements in order to describe upload speeds as fully 
as possible.
The following **graphs / tables** show these statistics for the sample as a whole,
as well as similar statistics related to frequency of testing.

@@

# Time series clustering

Clustering methods are a widely used family of techniques in **economic?** geography (**ADD REF**).
It can be defined within the modern machine learning framework as an unsupervised
learning task, which involves partitioning unlabeled observations into homogeneous
groups known as clusters [@montero2014tsclust]. 
The key idea is that observations within clusters tend to be more similar than 
observations between clusters.
Clustering is particularly useful for exploratory studies as it identifies
structures within the data [@aghabozorgi2015time].

Various applications of clustering can be found within geography: **ADD EXAMPLES**

Key characteristic of these studies is the cross-sectional nature of the data they
employed. 
Indeed, most clustering problems in geography deal with observations 
that are fixed in time. 
However, for this paper we are interested in creating clusters of local authorities 
in the UK with similar temporal signatures of experienced internet speeds over time. 
Hence, we deviate from the established geographical clustering tools and 
employ time series clustering methods.

Time-series is a common type of dynamic data that naturally arises in many 
different scenarios, such as stock data, medical data, and machine monitoring, 
just to name a few **(Aghabozorgi et al., 2015; Aggarwal and Reddy, 2013)**. 
They pose some challenging issues due to the large size and high dimensionality 
commonly associated with time-series **(Aghabozorgi et al., 2015)**. 
In this context, dimensionality of a series is related to time, 
and it can be understood as the length of the series.

Time series clustering methods have been developed in order to deal with clustering
problems linked to, for instance, stock or other financial data, economic,
governmental or medical data as well as machine monitoring
[@aggarwal2013time; @aggarwal2001surprising; @hyndman2015large; @WARRENLIAO20051857].
The main challenge -- and also the difference with cross-sectional clustering
problems -- is data dimensionality given the multiplicity of data points for every
individual included in the data set. 
Time series are dynamic data as the value of the observations change as 
a function of time [@aghabozorgi2015time].
This high dimensionality leads to 
(i) computational and algorithmic challenges regarding handling these data and 
building algorithms to perform clustering over long time series, 
and (ii) open questions regarding the choice of similarity measures in order to 
cluster similar times series together considering the whole length of the time series 
and overcoming issues around noise, outliers and shifts [@lin2004iterative; 
@aghabozorgi2015time].

Time series clustering methods utilising the whole length of time series can be
grouped in three categories. 
The first -- model-based approaches -- is based on recovering the underlying model 
for each time series and then apply clustering algorithms on the model parameters 
[@aghabozorgi2015time]. 
The main criticism is the cluster accuracy for near clusters [@mitsa2010temporal]. 
The second approach is a based the formation of vectors of features based on the 
original time series.
These new data of reduced dimensionality is then clustered using conventional
clustering algorithms.

For this paper we utilise the third category of time series clustering methods
known as shape-based approaches. The main idea is to match any two sperate time
series based on the similarity of their shapes through the calculation of
distances among these two shapes. Such approached are better equipped to capture
similarities between short length time series [@aghabozorgi2015time].
This approach serves best this paper because (i) we aim to identify clusters of
UK local authorities with similar temporal signatures -- i.e. shapes -- of
experienced internet speeds and (ii) the length of our time series is short
(**see also discussion in Section XX**).

Other important dimensions regarding time series clustering are related with the
clustering algorithm. Similar to clustering of cross-sectional data, we can employ
partitioning algorithms, which lead to non-overlapping clusters, hierarchical
clustering, which result to a hierarchy of clusters and fuzzy algorithms, which
create overlapping clusters [@sardatime]. Because of the simplicity of the
implementation and the interpretability of the results, we utilise here
partitioning clustering based on the widely *k-Means* algorithm. This is an iterative
algorithm, which begins with defining the desired number of clusters *k*. Then,
each observation is randomly assigned to a cluster from the $[1,k]$ space. This
is the initial cluster assignment, which is followed by iterations in order to
minimise the distance between the centroids of the clusters and the observations
assigned to these clusters [@james2013introduction].

There are a number of differences between the above described application of
*k-Means* for cross-sectional data and its application for times series data.

k-Meidoids
distance between points in time series
dt

comparison with density-based clustering DBSCAN [187]
where a cluster is expanded if its neighbours are dense.

- evaluation
- standardised

@sardatime
In the context of shape-based time-series clustering, it is common to utilize the Dynamic Time
Warping (DTW) distance as dissimilarity measure (Aghabozorgi et al., 2015). The calculation of the
DTW distance involves a dynamic programming algorithm that tries to find the optimum warping
path between two series under certain constraints

**moved this section from data and descriptives as need to end methods of clustering with how it applies to what we did, but needs further edit after you're done**
To apply these methods to our data, 
we aggregate all the speed-checks during the 13 weeks of March 
to May inclusive for weekdays in 2020 by each hour of the day and day of the week. 
As our research aims to identify the geography of internet service resilience for 
work purposes, bank holidays and the hours between midnight and 6am were also 
excluded. 
The composite week time series thus comprise 18 hours multiplied by 5 weekdays or 
90 time points per series. 
These time series were calculated for each of 
the 382 Local Authority Districts (LADs) in the UK, standardised, 
and then a k-means clustering algorithm was applied, including dynamic time warping. 
The LADs were assigned to 10 clusters for upload speeds. 
The cluster id for each LAD is then reattached to the speed-check dataframe to 
identify the characteristics of each cluster, including number of LADs, and the 
descriptives statistics of upload speeds in that cluster, 
and the temporal profile by hour of the day and day of the week.
Thus we aim to answer the first part of our research question: 
How resilient are broadband speeds as experienced in different parts of the UK during 
a time of extreme demand?

>>>>

---
title: |
  WFH and broadband speed (title needs rework)
type: ARTICLE TEMPLATE
author:
  - name: A. N. Author
    affil: a
    email: latex.helpdesk@tandf.co.uk
  - name: John Smith
    affil: b
    email: john.smith@uni-freiburg.de
affiliation:
  - num: a
    address: |
      Taylor & Francis, 4 Park Square, Milton Park, Abingdon, UK
  - num: b
    address: |
      Institut für Informatik, Albert-Ludwigs-Universität, Freiburg, Germany
bibliography: bibliography.bib
appendix: appendix.tex
abstract: |
  TBC
keywords: |
  covid; internet; working from home; broadband speed; time-series clusters
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

```{r settings, include = FALSE}
library(ggplot2)
library(tidyverse)
library(geojsonio)
library(tsfeatures)
library(rprojroot)
library(openair)
library(reshape2)
library(lubridate)
library(rgdal)
library(maptools)
library(sp)
library(rgeos)
library(classInt)
library(RColorBrewer)
library(knitr)
library(rprojroot)
library(viridis)
library(hrbrthemes)

# This is the project path
path <- find_rstudio_root_file()
```

@@

<!-- [our google doc](https://docs.google.com/document/d/1PWjkmgzWGYKR9wFogKYw7l-8mZLoORt593x-Tu-f2-M/edit#heading=h.i5om1o8wpcd9) -->

During the pandemic, working from home using digital technologies, whether partially
or exclusively, was transformed from a niche means of accessing work, albeit one that
had been on a slow, upward trend, to a widespread way of life in many countries.
The ability to work from home or telecommute meant millions retained their jobs and,
to a varying extent, maintained productivity during periods of strict
lockdown around the world.
However, this ability has not been evenly distributed socially or spatially,
creating a new type of digital divide.
On one side are those who can work from home, supported by digital technologies,
and have thus been able to enjoy both economic resilience and greater personal safety.
On the other side, previously employed individuals have been forced to accept
furlough or redundancy packages unless they are part of the cadre of essential
workers, who are potentially at high risk of infection.
Whilst the basis for this new digital divide has been viewed as mainly occupational,
here we consider whether the divide is also technological.

<!-- our approach -->
Using the UK as a case study, this paper aims to understand how the quality and
reliability of internet service, as reflected in *experienced* internet speeds,
may reinforce or redress the spatial and social dimensions of the digital division
exposed by the pandemic.
To do so, we employ volunteered geographic data on individual broadband speed tests
and state-of-the-art time-series clustering methods to create clusters of UK local
authorities with similar temporal signatures of experienced internet speeds.
We then associate these clusters of local authorities with their socioeconomic
and geographic characteristics to explore how they overlap with or diverge from
the existing economic and digital geography of the UK. **need to add urban-rural to analysis**
Our analysis enables us to better understand how the spatial and social distribution
of both occupations and online accessibility intersect to enable or hinder the
practice of telecommuting at a time of extreme demand.
We will also consider what lessons can be learned from this time for a future where
telecommuting is likely to remain a common means of accessing work, and broadband
services and infrastructure must be fit for purpose.
**LET'S LEAVE IT FOR NOW, BUT I THINK WE CAN CRYSTALISE MORE THE RQ**

<!-- capability to WFH -->
The capability to work from home has previously been studied from the
perspective of whether work tasks in a given occupation both can be and are
allowed to be performed using digital technologies independently of
location or co-location with colleagues, including supervisors
[@allen2015effective; @singh2013modeling].
However, successful telecommuting also requires that the quality and
reliability of digital services, particularly home internet connection speeds,
enable the completion of work tasks with a minimum of delay or interruption.
Prior to the pandemic, the performance of broadband services with respect to
telecommuters was never tested at scale, as working from home and connecting to
colleagues and workplace resources via the internet was the purview of a small minority of workers.
Instead, leisure use in the evening, when video streaming services are at their peak,
has been used to benchmark broadband performance and service delivery by different
Internet Service Providers (ISPs), at least in the UK [@ofcom2017].
Yet the shift towards telecommuting during various stages of lockdown around the world
has been drastic and there are speculations that post-Covid, the tendency to work from
home will be much higher than pre-Covid, raising questions around whether internet
services can accommodate the increased demand.
For example, $47$% of people in employment in the UK worked soley from home in April
$2020$, whilst the same figure only reached $5$% the year before [@ons2020;
@ons2020lm2019].
A back of the envelope calculation suggests that up to 40% of the working force
could work from home on an ongoing basis [@batty2020editorial].
Similar figures have been reported for other countries [@felstead2020homeworking].
For instance, $37$% of the European workforce worked from home in April $2020$ with
countries like Finland reaching $60$% [@eurofound2020].
In the US, almost half of the working population worked from home during the same
period because of the pandemic [@brynjolfsson2020covid],
and a recent estimate indicated that $37$% of all jobs in the US can be permanently
performed entirely from home [@NBERw26948].

<!-- ICT infrastructure and digital divides-->
None of these changes could have happened in the absence of reliable information
and communication technology (ICT) infrastructure -- both in terms of software
and hardware.
But while software innovations are easily diffused across space and
society^[See for example the huge success of videoconferencing apps such as Zoom
[@marks2020zoom].],
the same does not apply for ICT hardware infrastructure such as internet
broadband connectivity.
The literature describes first level digital divides in terms of the
availability and quality of internet connectivity, such as that manifest in
different geographies in the UK [@riddlesden2014broadband; @philip2017digital].
Second level digital divides consider the presence or lack of the necessary skills
to effectively utilise digital technologies and the internet
[@blank2014dimensions; @van2011internet].
The third level focuses on the heterogenous returns of internet usage among
different socioeconomic groups and, consequently, how digital technologies can
assist in bridging or further enhancing existing socioeconomic divides.
[@stern2009levels; @van2014digital; @van2015third].
The capability to telecommute is related to all three levels of digital divides,
but more importantly leads to differentiated outcomes regarding the economic
resilience of people and places to overcome a systemic shock such as the current
pandemic.
We identify this as a new digital divide, one that fundamentally alters the
potential returns of internet use for the user and wider community,
assumes skills or functions that are present in some occupations but not in others,
and relies upon access to high quality internet services.
**WE NEED TO CONNECT OUR FINDINGS WITH THIS AND JUSTIFY THAT THIS IS INDEED A NEW DIVIDE**
As the quality of internet infrastructure and services, as well as variation in
occupations are spatially dependent and clustered in space, our approach offers a
framework for understanding which types of places, are more likely to land on the
right side of this new digital divide.
By asking how resilient broadband speeds are as experienced in different parts of
the UK during a time of extreme demand, we interrogate which places benefit from
the greater economic resilience digital technologies can offer, not only during
the pandemic, but also into the future.

The structure of this paper is as follows.
First we review the literature on telecommuting and digital divides to better
understand the origins of the new digital divide revealed by the pandemic and its
impact on the economic resilience of different places.
We then describe our data and methodology.
Our results section first offers classification of how internet services vary
across the UK local authorities and then assesses whether these clusters replicate
or repudiate other socio-economic and geographic patterns of economic resilience.

@@

## From telecommuting to #WFH

@@

## Digital divides and economic resilience
<!-- WFH and digital divides-->
Our understanding of telecommuting as a product of enabled occupations can be
described as a manifestation of the third level digital divide, as those who are able
to use digital technologies to work from home benefit from a high rate of return on
their use of the internet in terms of autonomy, flexibility, and time saved from
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
These returns have been even greater during the Covid-19 crisis,
when those with the ability to telecommute also have the ability to maintain
their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the first
level digital divide, which is associated with access and quality of internet
connectivity.
For example, the systematic review from @SALEMINK2017360 highlights the
infrastructure quality differences between urban and rural areas
in various advanced economies.
Whether this variation in infrastructure quality affects the spatial footprint of
telecommuting has not previously been measured, although there are indications that
those who purchase high speed connections consume more data of all sorts and use their
connections for a variety of purposes [@hauge2011consumer].
Another study identifies a correlation between access to internet services and a
reduction in household transport spend across 33 countries [@bris2017ict].
Whether the implication is less travel because of increased telecommuting,
or whether internet access enables more efficient travel, this finding is
an indication of the potential household savings better internet services offer.
Thus, the extreme demand during the pandemic provides a new opportunity to
understand how infrastructure accessibility, quality, and reliability
affects telecommuting, particularly as working from home during the pandemic required
high volumes of bandwidth-intensive video conferencing in order to avoid the
face-to-face contact that could increase the spread of infection.
By first answering questions about internet service resilience, we can also refine our
understanding of how this has contributed to or reduced the new digital divide,
where economic resilience has been dependent upon the capability to work from home.

<!-- Covid-19 and working from home, cities, urban structure -->
The multi-layered digital divides intersect with materials divides and the economic
geography of the UK.
Following the regional economic resilience literature, which
underlines the differentiated capacity of cities and regions to escape or
recover from economic crises [@martin2012regional; @kitsos2018economic],
different places have different industrial and occupational profiles,
and these affect the aggregated potential capacity of places for telecommuting.
Such profiles are associated with longstanding inequalities in the UK and their
spatial representation as a North-South divide [@martin_north_south].
Various studies have illustrated severe inequalities between the north and the
south part of the UK regarding, for example, skills and human capital,
unemployment, productivity and prosperity [@lee2014grim; @mccann2020perceptions;
@dorling2018peak].
Some scholars have even argued that the UK suffers some of the highest level of
interregional inequalities in the global north [@gal2018reducing; @mccann2016uk].
Not only all three levels of digital divides are, to a certain extend, associated
with or shaped by the geography of the UK, but the intersection of the digital and
material divides affects the capacity of places to overcome, at least partially,
the economic effects of the Covid-19 pandemic.
Importantly, this is the first time that digital technologies became an essential
tool for economic resilience for such a great part of the population.

## other subsection?
**I THINK WE HAVE ENOUGH OF A LIT REV**
- circumstances where work can be carried out more flexibly in space and time
- how relate to digital divides
- capability theory of mobility?
- broadband tech stuff? broadband studies / divides / resilience

@@

# Methods and data

<!-- clustering in general -->
The starting point of our methodological framework is cluster analysis, which can
be defined within the modern machine learning framework as an unsupervised
learning task, which involves partitioning unlabelled observations into homogeneous
groups known as clusters [@montero2014tsclust].
The key idea is that observations within clusters tend to be more similar than
observations between clusters.
Clustering is particularly useful for exploratory studies as it identifies
structures within the data [@aghabozorgi2015time].
Therefore, cluster analysis is a widely used family of techniques in geography
[@gordon1977classification; @everitt1974cluster].
For instance, clustering methods is the basis of *geodemographics*, which aims
to create small area indicators or typologies of neighbourhoods based on various
and some times diverse variables [@SINGLETON2009289; @harris2005geodemographics].
Clustering techniques have also been employed to solve regionalisation problems
[@niesterowicz2016].

<!-- time-series clustering -->
Common characteristic of these studies is the cross-sectional nature of the data
they employ.
Indeed, most clustering problems in geography deal with observations that are
fixed in time.
However, for this paper we are interested in creating clusters of local authorities
in the UK with similar temporal signatures of experienced internet speeds over time.
Hence, we deviate from the established geographical clustering tools and
employ time-series clustering methods.

Time-series clustering methods have been developed in order to deal with clustering
problems linked to, for instance, stock or other financial data, economic,
governmental or medical data as well as machine monitoring
[@aggarwal2013time; @aggarwal2001surprising; @hyndman2015large; @WARRENLIAO20051857].
The main challenge -- and also the difference with cross-sectional clustering
problems -- is data dimensionality given the multiplicity of data points for every
individual object -- local authority in our case -- included in the data set.
Time-series are dynamic data as the value of the observations change as
a function of time [@aghabozorgi2015time].
This high dimensionality leads to
(i) computational and algorithmic challenges regarding handling these data and
building algorithms to perform clustering over long time-series,
and (ii) open questions regarding the choice of similarity measures in order to
cluster similar times series objects together considering the whole length of the time-series
and overcoming issues around noise, outliers and shifts [@lin2004iterative;
@aghabozorgi2015time].

Time-series clustering methods utilising the whole length of time-series can be
grouped in three categories.
The first -- model-based approaches -- is based on recovering the underlying model
for each time-series and then applying clustering algorithms on the model parameters
of each time-series [@aghabozorgi2015time].
The main criticism is the cluster accuracy for nearby clusters [@mitsa2010temporal].
The second approach is a based the formation of vectors of features based on the
original time-series.
These new data of reduced dimensionality is then clustered using conventional
clustering algorithms.

For this paper we utilise the third category of time-series clustering methods
known as shape-based approaches.
These methods match two sperate time-series objects based on the similarity of
their shapes through the calculation of distances among these two shapes and are
better equipped to capture similarities between short length
time-series [@aghabozorgi2015time].
This approach serves best this paper because (i) we aim to identify clusters of
UK local authorities with similar temporal signatures -- i.e. shapes -- of
experienced internet speeds and (ii) the length of our time-series is short
(see the data discussion in this section).

Another import element of time-series clustering is the actual clustering algorithm.
Similar to clustering of cross-sectional data, we can employ partitioning
algorithms, which lead to non-overlapping clusters, hierarchical
clustering, which result to a hierarchy of clusters and fuzzy algorithms, which
create overlapping clusters [@sardatime].
Because of the simplicity of the implementation and the interpretability of the
results, we utilise here partitioning clustering based on the widely used *k-Means*
algorithm.
This is an iterative algorithm, which begins with defining the desired number of
clusters *k*.
Then, each observation is randomly assigned to a cluster from the $[1,k]$ space.
This is the initial cluster assignment, which is followed by iterations in order
to minimise the distance between the centroids of the clusters and the observations
assigned to these clusters [@james2013introduction].

<!-- difference between clustering cross-sectional and time-series data -->
There are a number of differences between the above described application of
*k-Means* for cross-sectional data and its application for times series data.
Instead of creating clusters around centroids, a common approach is to create
clusters around *medoids*, which are representative time-series objects with a
minimal distance to all other cluster objects [@sardatime].
Also, instead of calculating the Euclidean distance between centroids and data
points, more complex distance measures need to be employed in order to capture
the similarity between a time-series object and a medoid.
A common distance measure for shape-based time-series clustering in the Dynamic
Time Warping (DTW).
Using its underpinning dynamic programming algorithm, DTW compares two time-series
objects to find the optimum warping path between them.
DTW is widely used in order to overcome limitations linked to the use of Euclidean
distance [@sardatime; @berndt1994using; @ratanamahatana2004everything].
The `R` package `dtwclust` has been used for the time-series clustering [@dtwclust].

<!-- introduction of our broadband speeds data -->
To assess the quality and reliability of internet across local authorities
in the UK during the time when the population were told to work from home if at
all possible we utilise unique data regarding individual internet speed tests from
Speedchecker Ltd^[[https://www.broadbandspeedchecker.co.uk/](https://www.broadbandspeedchecker.co.uk/)].
This a private company that allows internet users to check their own broadband
upload and download speeds, and stores every speed-check with timestamp and
geolocation information.
These data have been used before to assess digital divides [@riddlesden2014broadband]
and the impact of local loop unbundling regulatory processes [@nardotto2015unbundling].
These volunteered geographic data enable to assess the *experienced* internet
speed by users, which may differ from the *advertised* maximum speeds provided
by Internet Service Providers.
We are particularly interested in upload speeds and the frequency of speed tests.
While the former is less associated with internet-based leisure activities such
as video streaming, which involve downloading large amounts of data, the latter
can be linked more with work-related activities such as uploading documents or
two-way communication and interaction systems.
**ADD MORE. HOW ABOUT SPEED TESTS?**

<!-- data transformation -->
The first step in the workflow after dropping some outliers following
@riddlesden2014broadband is to transform and the individual,
geolocated and time-stamped tests to more meaningful aggregates both in terms
of space and time.
Regarding the temporal dimension, we aggregate all the speed-checks during the $13$
weeks of March to May inclusive for weekdays in $2020$ by each hour of the day and
day of the week.
As our research aims to identify the geography of internet service resilience for
work purposes, bank holidays and the hours between midnight and 6am were also
excluded.
The composite week time-series thus comprise 18 hours multiplied by 5 weekdays or
90 time points per series.
We aggregate these data because, we could not follow individuals or households
and connect data points.
Also, although this is a large data set ($241,088$ individual tests during the
working hours of study period), there are not enough observations for each local
and for each working hour ($631$ speed test per LAD in average).
These time-series were calculated for each of the 382 Local Authority Districts
(LADs) in the UK, standardised, and then a *k*-means partitioning around medoids
clustering algorithm was applied using DTW.
We initially run the algorithm for $k = 5, 10, 15 and 20$ and used cluster validity
indices (CVIs) to pick the optimal solution of $k = 10$.
Following @sardatime the following CVIs were employed: Silhouette (max), Score
function (max), Calinski-Harabasz (max), Davies-Bouldin (min), Modified Davies-Bouldin (DB*, min), Dunn (max), COP (min).

```{r echo=FALSE, eval=FALSE, message=FALSE}

# Code to calculate the above figures about N.

#upload data file
path.data <- paste(path, "./data/temp/TSbb19_20sp.csv", sep = "")
TSbb19_20sp <- read.csv(path.data)
TSbb20sp <- TSbb19_20sp %>%
  filter(dateOnly> "2019-12-31")

# N. of observations for study period  
print(nrow(TSbb20sp))

# observations for study period per local authority
print(nrow(TSbb20sp)/length(unique(TSbb20sp$LAD19NM)))
```

We then match the cluster membership to the LAD-level data to identify the
characteristics of each cluster, including number of LADs, the descriptive
statistics of upload speeds in that cluster, and the temporal profile by hour of
the day and day of the composite week.
Thus we aim to answer the first part of our research question:
How resilient are broadband speeds as experienced in different parts of the UK
during a time of extreme demand?

<!-- metrics -->
Since the quality and reliability of internet services vary in time and space due
to both supply and demand-side influences, we use a number of different measures
of experienced upload speeds.
These include: a) mean, experienced connection speed,
b) standard deviation or the amount of fluctuation from the mean, and
c) the variation in speeds at particular times of day when working from home is
more likely to take place.
We take account of all three measurements in order to describe upload speeds as
fully as possible.
The following **graphs / tables** show these statistics for the sample as a whole,
as well as similar statistics related to frequency of testing.

@@

<!-- HB old PARA4 - link to auxillary regressions method, or put in results to introduce results of auxillary regressions? -->
The cause of these different experiences of broadband resilience may be different in
different areas, as they may reflect either similarities in patterns of demand
or similar quality of infrastructure.
Our approach is also limited by potential endogeneity, as for example, better
quality connections with high mean speeds may enable more working from home, but
greater demand may cause slower speeds, less reliability and greater variability
of speed at different times of day or week.
Therefore, we avoid attributing any cause to our analysis of the experienced level
of quality and reliability of upload speeds.
Instead, we run an auxiliary regression in order to understand how the spatial and
temporal patterns of internet service relate to the economic geography of the UK.
We discuss how the different patterns might support or undermine efforts to work
from home and maintain safe productivity and whether they reinforce existing
spatial and social inequalities.
More specifically, we estimate the following multinomial model:

\begin{align}
Pr(Y_{i}=j) = \frac{exp^{X_{i}\beta_{j}}}{\sum_{i=1}^j exp^{X_{i}\beta_{j}}}
\begin{cases}
    i = 1, 2, ... , N \\t
    j = 1, 2, ... , J
\end{cases}\label{eq1}
\end{align}

The identification strategy is as follows.
Based on the outcomes of the time-series clustering, we identify $J$ distinct
and crisp clusters.
We then regress this cluster membership against a vector $X_{i}$ of
socio-economic and geographic variables, which we are discussed in details in
**section XX**.
This analysis enables us to provide a more nuanced understanding of how
telecommuting and technology intersect at a time of extreme demand, and what
lessons this time has for a future where telecommuting is likely to remain a
common means of accessing work and broadband services, as well as infrastructure,
must be fit for purpose.
